---
title: "SQL Project"
---

## Github Project Repo

[Github Repo](https://github.com/ADC-405-S26/Project_2_SQL_Team_7)

## Introduction

This project was focused on using SQL to query database data. My project partner Nick Katz and I focused on force plate jumping metrics. We created our own database using data from three different types of jumps. After created the database we used SQL code to make queries to extract data from the database to create visualizations to compare said data.

## The Data

These tables show the first few lines of the four datasets we combined into our database: Athlete data, CMJ jumps, MR jumps, SJ jumps.

```{r setup}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(readr)
library(knitr)
library(mdsr)
library(DBI)
library(RMariaDB)
library(dbplyr)
library(duckdb) 
```

```{r data loading and cleaning}
#| message: false
#| warning: false
#| include: false
#Loading - hawkinsr query for last 365 days of tests in KPI baseball players
CMJ_raw_data <- read.csv("sql_project/CMJ_Data.csv")
MR_raw_data  <- read.csv("sql_project/MultiRebound_Data.csv")
SJ_raw_data  <- read.csv("sql_project/SJ_Data.csv")

#Master athlete name / id file
athletes <- CMJ_raw_data %>%
  select(athlete_id, athlete_name) %>%
  distinct()
readr::write_csv(athletes, "athletes.csv")

#Cleaning: 257 -> 34-87 vars 
#In raw data, each dataset has columns for all jump types / tests. Each jump only has specific metrics and information, and this cleaning reduced all datasets to the jump-type specific variables, excluding all unrelated columns.

CMJ_clean_data <- CMJ_raw_data[-c(3:4,7:9,12:16,99:257)]
readr::write_csv(CMJ_clean_data, "CMJ_cleaned.csv")   

SJ_clean_data <- SJ_raw_data[-c(3:4, 7:9,12:16,20:23,26,29,31:34,36:37,41,43:44,46,51:54,59,62,65,67,69,73,75:79,81,85:87,90,92,95,98:99,101:257)]
readr::write_csv(SJ_clean_data, "SJ_cleaned.csv")

MR_clean_data <- MR_raw_data[-c(3:4,7:9,12:67,69:100,129:257)]
readr::write_csv(MR_clean_data, "MR_cleaned.csv")
```

```{r Database creation and optimization}
#| echo: false
#| message: false
#| warning: false
#| include: false
con_FP <- DBI::dbConnect(duckdb::duckdb(), dbdir = "sql_project/Foce_PlateDB")
```

```{sql connection=con_FP}
--| echo: false
SHOW TABLES;
DROP TABLE IF EXISTS CMJ_cleaned;
DROP TABLE IF EXISTS MR_cleaned;
DROP TABLE IF EXISTS SJ_cleaned;
DROP TABLE IF EXISTS athletes;
```

```{sql connection=con_FP}
--| echo: false
CREATE TABLE athletes (
athlete_id VARCHAR(255) PRIMARY KEY,
athlete_name VARCHAR(255) NOT NULL);
```

```{sql connection=con_FP}
--| echo: false
COPY athletes FROM 'athletes.csv' (FORMAT CSV, HEADER, NULL 'NA');
SELECT * FROM athletes LIMIT 5;
```

```{sql connection=con_FP}
--| echo: false
CREATE TABLE CMJ_cleaned (
id VARCHAR(255) PRIMARY KEY,
timestamp BIGINT,
testType_name VARCHAR(255),
testType_canonicalId VARCHAR(255),
athlete_id VARCHAR(255),
athlete_name VARCHAR(255),
right_avg_propulsive_force_n DOUBLE DEFAULT NULL,
relative_propulsive_net_impulse_n_s_kg DOUBLE DEFAULT NULL,
propulsive_phase_s DOUBLE DEFAULT NULL,
peak_relative_braking_force_percent DOUBLE DEFAULT NULL,
impulse_ratio DOUBLE DEFAULT NULL,
positive_impulse_n_s DOUBLE DEFAULT NULL,
countermovement_depth_m DOUBLE DEFAULT NULL,
left_force_at_peak_propulsive_force_n DOUBLE DEFAULT NULL,
jump_height_m DOUBLE DEFAULT NULL,
unweighting_phase_s DOUBLE DEFAULT NULL,
relative_propulsive_impulse_n_s_kg DOUBLE DEFAULT NULL,
landing_phase_s DOUBLE DEFAULT NULL,
l_r_avg_braking_rfd_percent DOUBLE DEFAULT NULL,
peak_relative_propulsive_power_w_kg DOUBLE DEFAULT NULL,
stiffness_n_m DOUBLE DEFAULT NULL,
avg_relative_braking_power_w_kg DOUBLE DEFAULT NULL,
unweighting_phase_percent DOUBLE DEFAULT NULL,
right_force_at_peak_braking_force_n DOUBLE DEFAULT NULL,
landing_performance_index DOUBLE DEFAULT NULL,
rsi DOUBLE DEFAULT NULL,
right_avg_braking_force_n DOUBLE DEFAULT NULL,
l_r_avg_propulsive_force_percent DOUBLE DEFAULT NULL,
avg_relative_propulsive_force_percent DOUBLE DEFAULT NULL,
left_avg_landing_force_n DOUBLE DEFAULT NULL,
l_r_avg_braking_force_percent DOUBLE DEFAULT NULL,
peak_propulsive_power_w DOUBLE DEFAULT NULL,
force_at_min_displacement_n DOUBLE DEFAULT NULL,
avg_braking_power_w DOUBLE DEFAULT NULL,
peak_landing_force_n DOUBLE DEFAULT NULL,
relative_braking_impulse_n_s_kg DOUBLE DEFAULT NULL,
right_avg_landing_force_n DOUBLE DEFAULT NULL,
avg_propulsive_force_n DOUBLE DEFAULT NULL,
flight_time_s DOUBLE DEFAULT NULL,
jump_momentum_kg_m_s DOUBLE DEFAULT NULL,
relative_braking_net_impulse_n_s_kg DOUBLE DEFAULT NULL,
peak_braking_force_n DOUBLE DEFAULT NULL,
avg_braking_velocity_m_s DOUBLE DEFAULT NULL,
left_avg_braking_force_n DOUBLE DEFAULT NULL,
landing_height_m DOUBLE DEFAULT NULL,
time_to_takeoff_s DOUBLE DEFAULT NULL,
avg_propulsive_velocity_m_s DOUBLE DEFAULT NULL,
takeoff_velocity_m_s DOUBLE DEFAULT NULL,
l_r_peak_braking_force_percent DOUBLE DEFAULT NULL,
time_to_stabilization_ms DOUBLE DEFAULT NULL,
avg_propulsive_power_w DOUBLE DEFAULT NULL,
right_avg_braking_rfd_n_s DOUBLE DEFAULT NULL,
l_r_peak_propulsive_force_percent DOUBLE DEFAULT NULL,
right_force_at_peak_propulsive_force_n DOUBLE DEFAULT NULL,
avg_relative_braking_force_percent DOUBLE DEFAULT NULL,
relative_peak_landing_force_percent DOUBLE DEFAULT NULL,
peak_relative_braking_power_w_kg DOUBLE DEFAULT NULL,
system_weight_n DOUBLE DEFAULT NULL,
m_rsi DOUBLE DEFAULT NULL,
peak_relative_propulsive_force_percent DOUBLE DEFAULT NULL,
peak_velocity_m_s DOUBLE DEFAULT NULL,
l_r_avg_landing_force_percent DOUBLE DEFAULT NULL,
positive_net_impulse_n_s DOUBLE DEFAULT NULL,
landing_stiffness_n_m DOUBLE DEFAULT NULL,
relative_force_at_min_displacement_percent DOUBLE DEFAULT NULL,
braking_net_impulse_n_s DOUBLE DEFAULT NULL,
braking_phase_percent DOUBLE DEFAULT NULL,
left_avg_braking_rfd_n_s DOUBLE DEFAULT NULL,
propulsive_phase_percent DOUBLE DEFAULT NULL,
avg_relative_propulsive_power_w_kg DOUBLE DEFAULT NULL,
braking_impulse_n_s DOUBLE DEFAULT NULL,
right_force_at_peak_landing_force_n DOUBLE DEFAULT NULL,
propulsive_net_impulse_n_s DOUBLE DEFAULT NULL,
l_r_peak_landing_force_percent DOUBLE DEFAULT NULL,
peak_braking_power_w DOUBLE DEFAULT NULL,
avg_braking_force_n DOUBLE DEFAULT NULL,
braking_phase_s DOUBLE DEFAULT NULL,
left_avg_propulsive_force_n DOUBLE DEFAULT NULL,
avg_landing_force_n DOUBLE DEFAULT NULL,
peak_braking_velocity_m_s DOUBLE DEFAULT NULL,
peak_propulsive_force_n DOUBLE DEFAULT NULL,
braking_rfd_n_s DOUBLE DEFAULT NULL,
left_force_at_peak_landing_force_n DOUBLE DEFAULT NULL,
l_r_propulsive_impulse_index_percent DOUBLE DEFAULT NULL,
left_force_at_peak_braking_force_n DOUBLE DEFAULT NULL,
l_r_landing_impulse_index_percent DOUBLE DEFAULT NULL,
propulsive_impulse_n_s DOUBLE DEFAULT NULL,
l_r_braking_impulse_index_percent DOUBLE DEFAULT NULL);

```

```{sql connection=con_FP}
--| echo: false
COPY CMJ_cleaned FROM 'CMJ_cleaned.csv' (FORMAT CSV, HEADER, NULL 'NA');
```

```{sql connection=con_FP}
--| echo: false
SHOW TABLES;
SELECT * FROM CMJ_cleaned LIMIT 5;
```

```{sql connection=con_FP}
--| echo: false
CREATE TABLE MR_cleaned (
id VARCHAR(255) PRIMARY KEY,
timestamp BIGINT,
testType_name VARCHAR(255),
testType_canonicalId VARCHAR(255),
athlete_id VARCHAR(255),
athlete_name VARCHAR(255),
system_weight_n DOUBLE DEFAULT NULL,
top_3_jumps_avg_jump_height_m DOUBLE DEFAULT NULL,
total_contact_time_s DOUBLE DEFAULT NULL,
top_5_jumps_avg_jump_height_m DOUBLE DEFAULT NULL,
top_5_jumps_avg_m_rsi DOUBLE DEFAULT NULL,
number_of_jumps_count DOUBLE DEFAULT NULL,
top_3_jumps_avg_m_rsi DOUBLE DEFAULT NULL,
top_5_jumps_peak_m_rsi DOUBLE DEFAULT NULL,
top_5_jumps_avg_rsi DOUBLE DEFAULT NULL,
l_r_avg_force_percent DOUBLE DEFAULT NULL,
top_3_jumps_peak_rsi DOUBLE DEFAULT NULL,
top_3_jumps_peak_m_rsi DOUBLE DEFAULT NULL,
top_5_jumps_peak_rsi DOUBLE DEFAULT NULL,
peak_jump_m_rsi DOUBLE DEFAULT NULL,
avg_contact_time_s DOUBLE DEFAULT NULL,
top_3_jumps_avg_rsi DOUBLE DEFAULT NULL,
peak_jump_rsi DOUBLE DEFAULT NULL,
avg_rsi DOUBLE DEFAULT NULL,
avg_jump_height_m DOUBLE DEFAULT NULL,
peak_force_n DOUBLE DEFAULT NULL,
avg_m_rsi DOUBLE DEFAULT NULL,
top_3_jumps_avg_contact_time_s DOUBLE DEFAULT NULL,
l_r_peak_force_percent DOUBLE DEFAULT NULL,
peak_jump_height_m DOUBLE DEFAULT NULL,
top_5_jumps_avg_contact_time_s DOUBLE DEFAULT NULL,
peak_rsi DOUBLE DEFAULT NULL,
total_flight_time_s DOUBLE DEFAULT NULL,
avg_force_n DOUBLE DEFAULT NULL,
peak_m_rsi DOUBLE DEFAULT NULL);

```

```{sql connection=con_FP}
--| echo: false
COPY MR_cleaned FROM 'MR_cleaned.csv' (FORMAT CSV, HEADER, NULL 'NA');
```


```{sql connection=con_FP}
--| echo: false
SHOW TABLES;
SELECT * FROM MR_cleaned LIMIT 5;
```

```{sql connection=con_FP}
--| echo: false
CREATE TABLE SJ_cleaned (
id VARCHAR(255) PRIMARY KEY,
timestamp BIGINT,
testType_name VARCHAR(255),
testType_canonicalId VARCHAR(255),
athlete_id VARCHAR(255),
athlete_name VARCHAR(255),
right_avg_propulsive_force_n DOUBLE DEFAULT NULL,
relative_propulsive_net_impulse_n_s_kg DOUBLE DEFAULT NULL,
propulsive_phase_s DOUBLE DEFAULT NULL,
left_force_at_peak_propulsive_force_n DOUBLE DEFAULT NULL,
jump_height_m DOUBLE DEFAULT NULL,
relative_propulsive_impulse_n_s_kg DOUBLE DEFAULT NULL,
landing_phase_s DOUBLE DEFAULT NULL,
peak_relative_propulsive_power_w_kg DOUBLE DEFAULT NULL,
landing_performance_index DOUBLE DEFAULT NULL,
l_r_avg_propulsive_force_percent DOUBLE DEFAULT NULL,
avg_relative_propulsive_force_percent DOUBLE DEFAULT NULL,
left_avg_landing_force_n DOUBLE DEFAULT NULL,
peak_propulsive_power_w DOUBLE DEFAULT NULL,
peak_landing_force_n DOUBLE DEFAULT NULL,
right_avg_landing_force_n DOUBLE DEFAULT NULL,
avg_propulsive_force_n DOUBLE DEFAULT NULL,
flight_time_s DOUBLE DEFAULT NULL,
jump_momentum_kg_m_s DOUBLE DEFAULT NULL,
landing_height_m DOUBLE DEFAULT NULL,
time_to_takeoff_s DOUBLE DEFAULT NULL,
avg_propulsive_velocity_m_s DOUBLE DEFAULT NULL,
takeoff_velocity_m_s DOUBLE DEFAULT NULL,
time_to_stabilization_ms DOUBLE DEFAULT NULL,
avg_propulsive_power_w DOUBLE DEFAULT NULL,
l_r_peak_propulsive_force_percent DOUBLE DEFAULT NULL,
right_force_at_peak_propulsive_force_n DOUBLE DEFAULT NULL,
relative_peak_landing_force_percent DOUBLE DEFAULT NULL,
system_weight_n DOUBLE DEFAULT NULL,
peak_relative_propulsive_force_percent DOUBLE DEFAULT NULL,
peak_velocity_m_s DOUBLE DEFAULT NULL,
l_r_avg_landing_force_percent DOUBLE DEFAULT NULL,
landing_stiffness_n_m DOUBLE DEFAULT NULL,
avg_relative_propulsive_power_w_kg DOUBLE DEFAULT NULL,
right_force_at_peak_landing_force_n DOUBLE DEFAULT NULL,
propulsive_net_impulse_n_s DOUBLE DEFAULT NULL,
l_r_peak_landing_force_percent DOUBLE DEFAULT NULL,
left_avg_propulsive_force_n DOUBLE DEFAULT NULL,
avg_landing_force_n DOUBLE DEFAULT NULL,
peak_propuslive_force_n DOUBLE DEFAULT NULL,
left_force_at_peak_landing_force_n DOUBLE DEFAULT NULL,
l_r_propulsive_impulse_index_percent DOUBLE DEFAULT NULL,
l_r_landing_impulse_index_percent DOUBLE DEFAULT NULL,
propulsive_impulse_n_s DOUBLE DEFAULT NULL,
propulsive_rfd_n_s DOUBLE DEFAULT NULL);
```

```{sql connection=con_FP}
--| echo: false
COPY SJ_cleaned FROM 'SJ_cleaned.csv' (FORMAT CSV, HEADER, NULL 'NA');
```

```{sql connection=con_FP}
--| echo: false
SHOW TABLES;
SELECT * FROM SJ_cleaned LIMIT 5;
```

## Querying the Data:

This query will take the left and right leg variables from both the CMJ and SJ tables and combine them to produce one dataset that can be used to compare the left and right legs. The dataset also has a metric (jump_height) to compare performance.

```{sql connection=con_FP}
--| output.var: "Left_Right_combined"
--| echo: false

SELECT a.athlete_name, a.athlete_id, sj.jump_height,
cmj.cmj_right_avg_propulsive_force_n, 
cmj.cmj_right_avg_landing_force_n, 
cmj.cmj_right_force_at_peak_propulsive_force_n, 
cmj.cmj_right_force_at_peak_landing_force_n, 

sj.sj_right_avg_propulsive_force_n,
sj.sj_right_avg_landing_force_n,
sj.sj_right_force_at_peak_landing_force_n,
sj.sj_right_force_at_peak_propulsive_force_n,

cmj.cmj_left_avg_propulsive_force_n, 
cmj.cmj_left_avg_landing_force_n, 
cmj.cmj_left_force_at_peak_propulsive_force_n, 
cmj.cmj_left_force_at_peak_landing_force_n, 

sj.sj_left_avg_propulsive_force_n,
sj.sj_left_avg_landing_force_n,
sj.sj_left_force_at_peak_landing_force_n,
sj.sj_left_force_at_peak_propulsive_force_n

FROM athletes AS a
LEFT JOIN (
  SELECT athlete_name, athlete_id,
    AVG(right_avg_propulsive_force_n) AS cmj_right_avg_propulsive_force_n, 
    AVG(right_avg_landing_force_n) AS cmj_right_avg_landing_force_n, 
    AVG(right_force_at_peak_propulsive_force_n) AS cmj_right_force_at_peak_propulsive_force_n, 
    AVG(right_force_at_peak_landing_force_n) AS cmj_right_force_at_peak_landing_force_n, 
    AVG(left_avg_propulsive_force_n) AS cmj_left_avg_propulsive_force_n, 
    AVG(left_avg_landing_force_n) AS cmj_left_avg_landing_force_n,
    AVG(left_force_at_peak_propulsive_force_n) AS cmj_left_force_at_peak_propulsive_force_n,
    AVG(left_force_at_peak_landing_force_n) AS cmj_left_force_at_peak_landing_force_n
  FROM CMJ_cleaned
  GROUP BY athlete_id, athlete_name)
AS cmj ON a.athlete_id = cmj.athlete_id
LEFT JOIN(
  SELECT athlete_name, athlete_id,
    AVG(right_avg_propulsive_force_n) AS sj_right_avg_propulsive_force_n, 
    AVG(right_avg_landing_force_n) AS sj_right_avg_landing_force_n, 
    AVG(right_force_at_peak_propulsive_force_n) AS sj_right_force_at_peak_propulsive_force_n, 
    AVG(right_force_at_peak_landing_force_n) AS sj_right_force_at_peak_landing_force_n, 
    AVG(left_avg_propulsive_force_n) AS sj_left_avg_propulsive_force_n, 
    AVG(left_avg_landing_force_n) AS sj_left_avg_landing_force_n,
    AVG(left_force_at_peak_propulsive_force_n) AS sj_left_force_at_peak_propulsive_force_n,
    AVG(left_force_at_peak_landing_force_n) AS sj_left_force_at_peak_landing_force_n,
    jump_height_m AS jump_height
  FROM SJ_cleaned
  GROUP BY athlete_id, athlete_name, jump_height)
AS sj ON a.athlete_id = sj.athlete_id
WHERE cmj.athlete_id IS NOT NULL AND sj.athlete_id IS NOT NULL;

```

```{r Checking second query}
head(Left_Right_combined)
```

## Data Visualization: Comparing Left and Right leg propulsive force

The first visualization shows the average propulsive force of the right leg of an athlete compared to the left leg. The second visualization shows the peak propulsive force of the right leg compared tot he left leg. As expected there is a strong correlation between the left and right leg for most athletes. Interestingly, there are several athletes who have both a great left and right leg propulsion, and they sit close tot he top on both, but for some reason they have a lower jump height, which doesn't make too much sense, as you would expect more force to equal more jump height. However, this doesn't take into consideration weight of the athlete when they are jumping, so heavier athletes may produce more force without jumping as high.

```{r}
#| echo: false
#| warning: false
library(tidyverse)

ggplot(data = Left_Right_combined) +
  geom_point(aes(x = sj_right_avg_propulsive_force_n, y = sj_left_avg_propulsive_force_n, color = jump_height), alpha = 0.5) + 
  labs(title = "Left and Right Leg Average Propulsive Force", 
       x = "Right Leg AveragePropuslive Force",
       y = "Left Leg Average Propuslive Force")

ggplot(data = Left_Right_combined) +
  geom_point(aes(x = sj_right_force_at_peak_propulsive_force_n, y = sj_left_force_at_peak_propulsive_force_n, color = jump_height), alpha = 0.5) + 
  labs(title = "Left and Right Leg Peak Propulsive Force", 
       x = "Right Leg Peak Propuslive Force",
       y = "Left Leg Peak Propuslive Force")

```

```{r}
#| echo: false
dbDisconnect(con_FP, shutdown = TRUE)
```

### ---------------------------------------------------------------------------

# Sources

* https://www.hawkindynamics.com/hawkin-metric-database

* https://www.hawkindynamics.com/blog/select-your-force-time-metrics-with-this-simple-system 

* https://www.sportsmith.co/articles/selecting-the-right-metrics-for-vertical-jump-testing/

* Google Gemini

### ---------------------------------------------------------------------------