---
title: "Text-Analysis Project"
---

## Github Project Files

https://github.com/ADC-405-S26/Project_3_TextAnalysis_Team_7

## Introduction & Background

Religious texts have shaped history not only through what they say, but also through how they are written. The four Gospels in the Bible are often read together because they each tell the story of Jesus from four different perspectives. While they overlap in many places, they also differ in structure, emphasis, and wording. These differences make the Gospels a useful case for text analysis, where we can compare not just what each Gospel says, but how each one says it. In this blog post, we will use computational tools to explore whether the Gospels’ language patterns reflect their historical and literary relationships. To be more specific, our primary goal is to identify which words and themes are most distinctive from one Gospel to the next, and understand the linguistic differences tied to the unique writing styles of each author.

For some brief context, historians generally agree that Mark was the first gospel to be written, around 70 A.D. Matthew and Luke were written between 80-85 A.D, and are generally believed to have used Mark as a source. These three are known as the synoptic gospels, a term derived from the idea of being “seen together. There are many similarities between these three writings, as they share many of the same stories, structures and, as we see, even wording patterns. Our cluster analysis attempts to quantitatively demonstrate this well-understood belief. 

Apart from these three, there is John. His gospel is believed to have been written last, around 90 A.D. There are many differences between John and the synoptic gospels, for instance, there are several stories told in John that do not appear in the other three, and parts of the chronology are different. 


```{r}
#| include: false

library(tidyverse)
library(scales)
library(tidytext)
library(textdata)
library(gutenbergr)
library(stylo)
library(quanteda)
library(wesanderson)
library(scales)
library(zoo)

 # install.packages("tidyverse")
 # install.packages("scales")
 # install.packages("tidytext")
 # install.packages("textdata")
 # install.packages("gutenbergr")
 # install.packages("stylo")
 # install.packages("quanteda")
 # install.packages("wesanderson")
 # install.packages("zoo")
```


## Data Explanation

The data used in this project contains the King James version of the four Gospels: Matthew, Mark, Luke, and John, from Gutenberg metadata. Each observation represents a portion of Gospel text, organized by Gospel and verse-level information. The raw version of our initial dataset includes the Gutenberg-assigned ID for each gospel, the free-form text, the title, and the Gospel name. 

```{r}
#| include: false
#| fig-align: "center"

data("gutenberg_metadata")

kingJamesBible_all <- gutenberg_metadata |>
  filter(str_detect(title, "Bible"))
# accessing data from the gutenberg library, checking for version of bible
# The King James Bible

Matthew <- gutenberg_download(8040, meta_fields = "title")
Mark <- gutenberg_download(8041, meta_fields = "title")
Luke <- gutenberg_download(8042, meta_fields = "title")
John <- gutenberg_download(8043, meta_fields = "title")
# getting the exact ID for the entry


```


```{r}
#| include: false

# downloading files into one dataset

gospel_ids <- tibble(
  gutenberg_id = c(8040, 8041, 8042, 8043),
  gospel = c("Matthew", "Mark", "Luke", "John")
)

gospels_raw <- gutenberg_download(
  gospel_ids$gutenberg_id,
  meta_fields = "title"
) |>
  left_join(gospel_ids, by = "gutenberg_id")

glimpse(gospels_raw)

```

```{r}
#| echo: false

head(gospels_raw)
```


```{r}
#| include: false

# cleaning textual data

gospels_clean <- gospels_raw |>
  filter(text != "") |>   # removing any rows where the text is now blank.

  # removing lines that begin with "book" followed by a space.
  # str_detect() checks whether a text pattern is present.
  filter(!str_detect(text, regex("^book\\s+", ignore_case = TRUE))) |>
  mutate(
    # str_replace_all() replaces every match of a pattern.
    # \\d+ - "one or more digits."
    # this removes chapter numbers, verse numbers, or labels
    text = str_replace_all(text, "\\d+", " "),
    
    # removing leftover colons from the beginning of the line
    # \\s* - "zero or more spaces"
    # : - "match a colon"
    # :? - "optionally matches one more colon"
    # the "" at the end deletes the matched punctuation
    text = str_replace(text, "^\\s*:\\s*:?\\s*", ""), 
    
    text = str_squish(text) # str_squish() removes extra spaces from each line of text.
  )
```


## Rolling Sentiment Analysis of each Gospel

```{r}
#| include: false

gospel_words <- gospels_clean |>
  tidytext::unnest_tokens(output = word, input = text, token = "words") |> 
  filter(!(word %in% stop_words$word)) |> # filtering out the non(!) stop words
  filter(!(word %in% "ye"), !(word %in% "thou"), !(word %in% "thy"), !(word %in% "thee"))

gospel_words |> 
  group_by(gospel) |>
  filter(word == "god") |>
  count(word) |>
  arrange(desc(n))
```


First, we can look at an overall sentiment analysis of each of the four gospels. A sentiment
analysis takes a look at the vocabulary used throughout a text and determines how positive or
negative the text as a whole is. The sentiment of each word is placed on a scale from -5 to
5, where a negative number denotes more negative sentiment, while a positive number denotes
positive sentiment. Before assigning a value to each word in the text, we first removed simple
words that don’t have a sentiment, words such as: and, thee, thou, and ye to name a few.
We determined that a rolling sentiment analysis would be most appropriate for this task, so
each data point on the plot takes into account the mean sentiment for the last 100 words.
This provides a much smoother way to visualize how the sentiment changes as the narrative
progresses. We also normalized each graph so that each of the plots are the same length.


```{r}
#| echo: false
#| message: false
#| warning: false
#| results: hide
#| fig-align: "center"


gospel_words <- gospels_clean |>
  unnest_tokens(word, text)

afinn <- tidytext::get_sentiments("afinn") # attaches a scale from -5 (negative) to +5 (most positive)

words_sentiment <- gospel_words |> # joins sentiment value with gospels
  inner_join(afinn, by = "word")

words_sentiment <- words_sentiment |> 
  group_by(gospel) |>
  mutate(index = row_number()) |> # index the rows to create a chronological order
  ungroup()

words_sentiment <- words_sentiment |>
  group_by(gospel) |>
  mutate(rolling_sentiment = rollmean(value, k = 100, fill = NA)) |> # create a rolling sentiment analysis
  ungroup()

words_sentiment <- words_sentiment %>%
  group_by(gospel) %>%
  mutate(norm_index = index / max(index)) %>% # normalize the length of each graph
  ungroup()

# creates the plot
ggplot(words_sentiment, aes(x = norm_index, y = rolling_sentiment)) + 
  geom_line(color = "steelblue") +
  facet_wrap(~gospel, scales = "free_x") +
  labs(
    title = "Sentiment Over Narrative by Gospel",
    x = "Narrative Progression",
    y = "Sentiment (Smoothed)") +
  theme_grey() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"))
    

```

We can see that there are stark difference between the narratives. Luke and Matthew have somewhat similar narrative progressions due to the fact that they were written around the same time, and likely used the same sources. Mark has the most muted account, where we see that the sentiment is not as extreme as the others. This makes sense because Mark is a much more blunt account of Christs life and chooses to focus on His deeds and teachings without too much filler.

We can observe that Mathew, Luke, John all have declining narratives, and then a sharp increase in sentiment a the end. This is due to the portrayal of Christs death and resurrection, the former obviously being negative and the latter being positive. However, this increase is noticeably absent from Marks account. As previously mentioned Mark is the shortest of all the accounts, and comes to a fairly abrupt end after Christs resurrection. Therefore, we don't see the same jump in sentiment that we observe in the other accounts. Overall, this sentiment analysis allows us to perceive narrative progression throughout the text, and gives us insight into how each author structured their respective narrative. 

```{r}
#| include: false

# manually creating themes for different words we want to focus on for tf-idf analysis
# also, because many of these words were not assigned a score through sentiment analysis,
# manually assigning range from -2 to 2
focus_words <- tribble(
  ~word, ~connotation, ~theme,
  
  # theme forgiveness
  "forgive", 2, "forgiveness",
  "forgiven", 2, "forgiveness",
  "repent", 1, "forgiveness",
  "repentance", 1, "forgiveness",
  
  # theme love
  "love", 2, "love",
  "loved", 2, "love",
  "loveth", 2, "love",
  "mercy", 2, "love",
  "merciful", 2, "love",
  "compassion", 2, "love",
  "pity", 1, "love",
  "grace", 2, "love",
  "way", 1, "love",
  "life", 0, "love",
  "light", 2, "love",
  
  # theme miracles
  "miracle", 2, "miracles",
  "miracles", 2, "miracles",
  "sign", 1, "miracles",
  "signs", 1, "miracles",
  "heal", 2, "miracles",
  "healed", 2, "miracles",
  "healing", 2, "miracles",
  "sick", 0, "miracles",
  "blind", -1, "miracles",
  "lame", -1, "miracles",
  "leper", 0, "miracles",
  "mighty works", 2, "miracles",
  "wonders", 2, "miracles",
  
  # theme evil
  "devils", -2, "evil",
  "devil", -2, "evil",
  "demon", -2, "evil",
  "demons", -2, "evil",
  "unclean", -2, "evil",
  "darkness", -2, "evil",
  "tempted", -2, "evil",
  "betray", -2, "evil",
  "betrayed", -2, "evil",
  "accuse", -2, "evil",
  "accused", -2, "evil",
  "enemy", -2, "evil",
  "enemies", -2, "evil",
  "pharisees", 0, "evil",
  "scribes", 0, "evil",
  "sin", -2, "evil",
  "sins", -2, "evil",
  "sinner", -2, "evil",
  "sinners", -2, "evil",
  "judgment", -2, "evil",
  "condemn", -2, "evil",
  "condemned", -2, "evil",
  
  # theme faith
  "faith", 2, "faith",
  "believe", 2, "faith",
  "believed", 2, "faith",
  "believeth", 2, "faith",
  "belief", 2, "faith",
  "witness", 1, "faith",
  "testimony", 1, "faith",
  "testify", 1, "faith",
  "truth", 2, "faith",
  "abide", 1, "faith",
  "dwelt", 1, "faith",
  "spirit", 0, "faith",
  "risen", 1, "faith",
  "rise", 1, "faith",
  "resurrection", 2, "faith",
  "tomb", 0, "faith",
  "reborn", 1, "faith",
  "born again", 1, "faith",
  "glory", 2, "faith",
  "jesus", 1, "faith",
  "son of god", 1, "faith",
  
  # theme death
  "death", -2, "death",
  "die", -2, "death",
  "died", -2, "death",
  "kill", -2, "death",
  "killed", -2, "death",
  "crucify", -2, "death",
  "crucified", -2, "death",
  "cross", -2, "death",
  "buried", -1, "death",
  "blood", -1, "death",
  

) |>
  distinct(word, .keep_all = TRUE)
```


```{r}
#| include: false

# cleaning focus_word df
focus_words_clean <- focus_words |>
  mutate(
    
    # str_replace_all() here replaces spaces with underscores for phrase matching
    # i.e.: "son of god" --> "son_of_god"
    feature = str_replace_all(word, "\\s+", "_")) |>
  
  # keeping only one row per feature
  distinct(feature, .keep_all = TRUE)


# pulling out only the multi-word phrases from the focus words
# str_detect here checks whether the word contains a space
# pull(word) then turns that column into a vector
multiword_terms <- focus_words_clean |>
  filter(str_detect(word, "\\s+")) |>
  pull(word)
```

```{r}
#| include: false

# collapsing the verses into one document per Gospel

gospel_docs <- gospels_clean |>
  group_by(gospel) |>
  summarize(
    
    # str_c() combines all verse text from each Gospel into one long string
    text = str_c(text, collapse = " "),
    .groups = "drop")

# creating quanteda corpus object for storing the texts
gospel_corpus <- corpus(
  gospel_docs,
  text_field = "text")

docnames(gospel_corpus) <- gospel_docs$gospel

# tokenizing the text
gospel_tokens <- gospel_corpus |>
  tokens(
    remove_punct = TRUE, # removes punctuation
    remove_numbers = TRUE,  # removes numbers
    remove_symbols = TRUE   # removes symbols
  )

# combining the multi-word phrases into single tokens
# this is necessary so multiple word phrases aren't counted separately
# tokens_compound() keeps selected phrases together
# clarifying the 'concatenator' connecting these phrases
if (length(multiword_terms) > 0) {
  gospel_tokens <- gospel_tokens |>
    tokens_compound(
      pattern = phrase(multiword_terms),
      concatenator = "_")}

# creating document-feature matrix, necessary for using quanteda
# this tell us how many times each feature appears in each Gospel
# dfm() creates the matrix
gospel_dfm <- dfm(gospel_tokens)

# now filtering out only the previously determined focus words
# dfm_select() filters the matrix keeping only this
focus_dfm <- dfm_select(
  gospel_dfm,
  pattern = focus_words_clean$feature,
  selection = "keep",
  valuetype = "fixed")

# dfm_tfidf() calculates the raw tokens into tf-idf scores
# scheme_tf = "prop" uses the proportional word frequency (accounts for different Gospel lengths)
focus_tfidf <- dfm_tfidf(
  focus_dfm,
  scheme_tf = "prop")

```


```{r}
#| include: false

# converting the TF-IDF matrix into long format for plotting purposes
focus_tfidf_tidy <- as.data.frame(
  as.matrix(focus_tfidf),
  # check.names = FALSE keeps R from changing feature names
  check.names = FALSE) |>
  
  # rownames_to_column() turns Gospel names into a real column
  rownames_to_column("document") |>
  
  # converting from wide to long format 
  pivot_longer(
    cols = -document,
    names_to = "feature",
    values_to = "tfidf") |>
  
  # rename() changes document to gospel
  rename(gospel = document) |>
  
  # joining the tf-idf results back to focus word df
  left_join(
    focus_words_clean,
    by = "feature") |>
  
  # calculating the weighted contribution score of each feature
  # this score represents both how important the word is to the specific Gospel,
  # and how strongly positive or negative (sentiment, not numeric) it is
  mutate(
    word_contribution = tfidf * connotation)
```

```{r}
#| include: false

# checking to see tf-idf scores against and calculated metric
focus_tfidf_tidy |>
  filter(tfidf > 0) |>
  arrange(desc(tfidf)) |>
  head(25)

# shows which words were not included in Gospels form selected focus words
setdiff(focus_words_clean$feature, featnames(focus_dfm))

```
 

```{r}
#| include: false

# because some themes contain more words than others, counting the number of terms 
# in each theme to scale them and create one normalized score
theme_sizes <- focus_words_clean |>
  count(theme, name = "number_of_terms")

# creating one score for each Gospel and theme combination
theme_scores <- focus_tfidf_tidy |>
  group_by(gospel, theme) |>
  summarize(
    
    # total_tfidf shows how distinctive the theme is,
    # and ignores whether the connotation is positive or negative
    total_tfidf = sum(tfidf),
    
    # raw_theme_score adds up tf-idf × connotation for all words in the theme
    # positive theme words increase score, vice-versa
    raw_theme_score = sum(word_contribution),
    .groups = "drop") |>
  
   # Add the number of focus terms in each them
  left_join(theme_sizes, by = "theme") |>
  
   # now normalizing the score by dividing by the number of terms in the theme
   # this will be used in heatmap
  mutate(
    normalized_theme_score = raw_theme_score / number_of_terms)
   

head(theme_scores)

```

## Exploring Distinctive Thematic Language between the four Gospels

First, we wanted to compare the Gospels through the specific types of language they emphasize. To do so, we split core words and phrases throughout the bible into 6 themes for later sentiment analysis: forgiveness, love, faith, miracles, evil, and death. Focusing on a smaller subset of words forces the analysis to center more directly on the theological and emotional language that shapes each Gospel. However, using sentiment analysis with religious phrases and wording can become quite buggy. Therefore, we had to manually create a 'lexicon' that not only categorized the custom set of words into these themes, but also gave a connotation score from -2 to 2: negative values represent darker or conflict-oriented language and positive values represent redemptive or hopeful language.

Since TF-IDF only shows which words are most distinctive in each Gospel, we combined it with our connotation scores to make the measure more interpretive. In other words, a weighted connotation score was created by multiplying each word’s TF-IDF value by its manually assigned connotation score. This means that a word only has a strong impact if it is both distinctive to a Gospel and strongly positive or negative in the lexicon. A positive word with a high TF-IDF score increases the theme’s overall score, while a negative word with a high TF-IDF score decreases it. Words with low TF-IDF values contribute less, even if they have a strong connotation, because they are not especially distinctive to that Gospel.

Below is a sample table showing how a few focus words contribute to the comparison between the four Gospels.
```{r}
#| echo: false

# THIS IS NOT NEEDED FOR ACTUAL HEATMAP

# small table for demonstration in final pdf
focus_tfidf_tidy |>
  filter(tfidf > 0) |>
  arrange(desc(tfidf)) |>
  select(
    Gospel = gospel,
    Word = word,
    Theme = theme,
    Connotation = connotation,
    `TF-IDF` = tfidf,
    `Weighted contribution` = word_contribution) |>
  
  # showing only 4 decimals
  mutate(
    `TF-IDF` = round(`TF-IDF`, 4),
    `Weighted contribution` = round(`Weighted contribution`, 4)) |>
  head(10) |>
  
  # knitr::kable() is better for formatting the table so it looks cleaner in final render
  knitr::kable(
    caption = "Sample of focus words after converting TF-IDF output into tidy format")

```

Now we visualized these scores using a heatmap, with the x-axis comparing Matthew, Mark, Luke, and John, and the y-axis listing the six themes. Each tile represents one Gospel-theme combination, with the color and number showing whether that theme is more positively or negatively weighted in that Gospel. Strong positive values indicate that a theme is being shaped by words that are both distinctive to that Gospel and positively coded. Strong negative values indicate that a theme is being shaped by words that are both distinctive to that Gospel and negatively coded. These values are the normalized theme scores multiplied by 1000, which was done only to make the very small TF-IDF-based values easier to read on the plot.

---

```{r, fig.width=9, fig.height=6}
#| echo: false
#| message: false
#| warning: false
#| results: hide
#| fig-align: "center"

# testing palettes
rocket_colors <- wes_palette("BottleRocket2", n = 4)
cavalcanti_colors <- wes_palette("Cavalcanti1", n = 4)

# adjusting numeric score for readability purposes
theme_scores_display <- theme_scores |>
  mutate(
    display_score = normalized_theme_score * 1000)

# making the color scale for heatmap (balanced around 0)
max_abs_score <- max(abs(theme_scores_display$display_score), na.rm = TRUE)

ggplot(theme_scores_display,
  aes(gospel, theme, fill = display_score)) +
  
  geom_tile(color = "white") +
  
  geom_text(
    aes(label = number(display_score, accuracy = 0.1)),
    size = 3.5) +
  
  scale_fill_gradient2(
    low = "brown1",
    mid = "white",
    high = "cyan3",
    midpoint = 0,
    limits = c(-max_abs_score, max_abs_score),
    labels = number_format(accuracy = 0.1),
    name = "Score\n×1000") +
  
  labs(
    title = "Thematic TF-IDF x Connotation Scores Across the Four Gospels",
    subtitle = "Scores combine focus-word distinctiveness with positive or negative connotation ratings",
    x = "Gospel",
    y = "Theme") +
  
  theme_void() +
  theme(
    plot.background = element_rect(color = "#2c3e50", linewidth = 0.4),
    plot.margin = margin(t = 8, r = 8, b = 8, l = 8),
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 11, face = "bold"))

```

Luke, Mark, and Matthew all demonstrate strong similarities in the forgiveness, miracles, and faith themes. Forgiveness appears to be heavily significant across these three Gospels, with Luke having the highest weighted connotation score at 2.3, followed by Mark at 1.8 and Matthew at 1.7. In contrast, the textual data in John results in a value of 0. This suggests that, based on the forgiveness words included in the lexicon, the Gospel of John did not contain any matches for this theme. Or, for this comparison, the theme of forgiveness is less evident than the other Gospels. Still, of course, this does not inherently mean that the concept of forgiveness is absent from John; rather, it means that this specific method of analysis did not identify words from that theme in the text. At the same time, John also shows stronger results in the themes of miracles and faith. Its weighted connotation scores for these themes are higher than the other Gospels, indicating that John may place greater emphasis on stories of miracles and faith than the other Gospels. Otherwise, the themes of love, evil, and death remain quite similar between Gospels.


## Using 'stylo' R Package to Measure Literary Style

To further analyze stylometric (the computational analysis of literary style) patterns in the four Gospels, we used the stylo R package. With this package, we wanted to better identify the distinctive writing styles of Matthew, Mark, Luke, and John, as well as understand the unique linguistic patterns among them. This is done through ‘stylos’ processing procedures, which measure features such as word frequency and vocabulary. In machine-learning terms, this is an unsupervised analysis, as the algorithm is not trained to predict any known variable but instead groups the Gospels based on their stylistic similarities and differences. For this analysis, we used the King James Version texts from Project Gutenberg and compared the four texts specifically using the 300 most frequent words.

Therefore, 'stylo' essentially compares how often words are repeated across the texts and then calculates which Gospels are most similar or most different from one another. We should also clarify that, because our analysis uses an English translation, the results should be interpreted as patterns specific to the KJV version of the Gospels and may not generalize to other versions.  
```{r}
#| include: false

# making the cleaned Gospel text for stylo package, combines all rows for each Gospel into one full document

gospels_for_stylo <- gospels_clean |>
  group_by(gospel) |>
  summarize(
    full_text = paste(text, collapse = " "), # collapse = " " adds a space between lines so words do not run together
    .groups = "drop"
  )

```


```{r}
#| include: false

# creating a folder named "corpus" where the text files for stylo can be stored and later used
dir.create("corpus", showWarnings = FALSE)

# this just deletes any old .txt files already inside the corpus folder (for rerunning the code)
unlink(list.files("corpus", pattern = "\\.txt$", full.names = TRUE))
```

```{r}
#| include: false

# creating a folder/file path for each Gospel text
gospels_for_stylo |>
  mutate(
    # paste0() combines text together with no spaces
    # this then creates files for each individual gospel
    file_name = paste0("corpus/", gospel, "_KJV.txt")
  ) |>
  # pwalk() goes row by row and uses the values from each column
  pwalk(function(gospel, full_text, file_name) {
    # write_lines() saves the full Gospel text into the file name we created
    write_lines(full_text, file_name)
  })

# checking that the Gospel text files were successfully created in the corpus folder
list.files("corpus")
```

```{r}
#| include: false

# creating a reusable function to run the stylo analysis
# this just helps prevent us from rewriting the same stylo() code multiple times
run_stylo <- function(mfw_number, analysis_type) {
  
  # stylo() compares writing style across the text files in the corpus folder
  stylo(
    gui = FALSE,              # this runs stylo without opening the interactive menu
    corpus.dir = "corpus",    # tells stylo where the Gospel text files are stored
    mfw.min = mfw_number,     # setting the minimum number of most frequent words to analyze
    mfw.max = mfw_number,     # setting the maximum number of most frequent words
    mfw.incr = 100,           # controlling how stylo increases word counts across runs
    analysis.type = analysis_type # telling stylo what kind of analysis to run
  )
}
```

```{r}
#| include: false


# Choosing 300 for most-frequent-words setting in stylo analyses
selected_mfw <- 300

```



```{r}
#| echo: false
#| message: false
#| warning: false
#| results: hide
#| fig-width: 8
#| fig-height: 5
#| out-width: "90%"
#| fig-align: "center"

# setting base R plot options for this stylo plot
# bg = "grey96" gives the plot a light gray background
# mar controls the plot margins in this order
# bottom, left, top, right.
par(
  bg = "grey96",
  mar = c(3, 3, 2, 2)
)

# running cluster analysis using stylo.
# this compares the Gospels based on the selected number of most frequent words
stylo_ca_300 <- stylo(
  gui = FALSE,
  corpus.dir = "corpus",
  mfw.min = selected_mfw,
  mfw.max = selected_mfw,
  mfw.incr = 100,
  analysis.type = "CA",

  # removing stylo's automatic titles to keep the plot cleaner
  titles.on.graphs = FALSE,
  custom.graph.title = "",

  # using stylo's automatic colors for the Gospel labels.
  colors.on.graphs = "colors",

  # Adjust the plot size for readability
  plot.custom.width = 8,
  plot.custom.height = 5,
  plot.font.size = 11,
  plot.line.thickness = 2,

  # displaying the dendrogram horizontally
  dendrogram.layout.horizontal = TRUE)

grid(
  col = "grey",
  lty = "dotted",
  lwd = 0.8)

# Add a darker border around the plot.
box(lwd = 2, col = "grey25")
title("CA Plot of the Gospels Using 300 Most Frequent Words")

```

This visualization is a cluster analysis plot, which works like a family tree of textual similarity. Texts that connect on shorter branches are more similar to one another, while texts that connect farther away are more distant in text patterning. To be more specific, the x-axis shows relative Euclidean distances, which represent stylistic similarity (lower values indicate greater similarity; higher values indicate greater dissimilarity). Matthew and Luke both sit at roughly 0.8, showing that they are the closest pair in this clustering. John’s Gospel seems to be at a much larger distance, demonstrating that it is the most linguistically distinct book in the analysis. additionally, as we have seen in historical context, the timeline of these words are evident in the cluster analysis, where John sits far from the other three. 

---

Overall, our analyses show that computational text analysis can reveal meaningful differences and similarities between the four Gospels that align with their historical and literary relationships. These findings show that while the four Gospels share many central religious themes, they differ in how strongly they emphasize certain ideas and in the structure of their language and wording. These differences highlight the distinct perspectives each Gospel brings to the larger narrative of Jesus’ life and teachings.






## References/Sources

* https://www.britannica.com/topic/biblical-literature/The-Synoptic-Gospels?utm_source

* https://www.bartehrman.com/messianic-secret/